Note on Linear Biases in MORUS

### 2018-04-13 morus.tex
Proofreading

\usepackage{tikz}
\usetikzlibrary{calc,math}

@@

In order to build the equation for the full trail, we start with $E^2_0$:
\[
S^2_{2,0} \oplus S^{3}_{2,b_2} \oplus S^{3}_{0,0} = S^2_{3,0} \cdot S^2_{4,0}.
\]
In order to cancel the $S^{3}_{0,0}$ term on the left-hand side, we add in the equation $A^2_{-b_0}$ (where the sum of two equations of the form $a = b$ and $c = d$ is defined to be $a+c = b+d$). This yields:
\begin{align*}
&S^2_{2,0} \oplus S^3_{2,b_2} \oplus C^2_{-b_0}\\
=\; &S^2_{3,0} \cdot S^2_{4,0} \oplus S^2_{1,-b_0} \oplus S^2_{3,-b_0} \oplus S^2_{1,-b_0} \cdot S^2_{2,-b_0} \oplus S^2_{2,-b_0} \cdot S^2_{3,-b_0}.
\end{align*}
We then need to cancel two terms of the form $S^t_{2,i}$. To do this, we add in the equations $D^t_i$ for appropriate choices of $t$ and $i$. This replaces the two $S^t_{2,i}$ terms by four $S^t_{4,i}$ terms. By using equation $B^t_i$ four times, we can then replace these four $S^t_{4,i}$ terms by eight $S^t_{1,i}$ terms. By applying equation $B^t_i$ eight times, these eight $S^t_{1,i}$ terms can in turn be replaced by eight $S^t_{0,i}$ terms (and some ciphertext terms). Finally, applying $A^t_i$ eight times allows us to replace these eight $S^t_{0,i}$ terms by only ciphertext bits. Ultimately, for \cipher{MORUS640}, this yields the equation:

@@

The equation for \cipher{MORUS1280} is very similar, and is given in \Cref{sec:traileq}.

@@

In the equation for \cipher{MORUS640} from the previous section, each line on the right-hand side of the equality involves distinct $S^t_{i,j}$ terms, and each line has a weight of 1. By the piling-up lemma, it follows that if we assume distinct $S^t_{i,j}$ terms to be uniform and independent, then the expression on the right-hand side has a weight of 16. Hence the linear combination of ciphertext bits on the left-hand side has an expected bias of $2^{-16}$. The same holds for \cipher{MORUS1280} (cf. \Cref{sec:traileq}).

@@

\begin{table}[h!]
  \caption{Experimental verification of trail biases.}
  \label{tab:miniapproximations}
  \centerline{
  \begin{tabular}{@{}llSSS@{}}
    \toprule
    & & \multicolumn{3}{@{}c@{}}{Weight} \\
    \cmidrule{3-5}
    \multicolumn{2}{@{}l}{Approximations for \cipher{MiniMORUS640}}
             & {Exp.} & {Bool.} & {Meas.} \\
    \midrule
    $\chi_1$ & $S^{2,2}_0 = C^0_{27} \oplus C^1_{0, 8, 26} \oplus C^2_{7,13,31} \oplus C^3_{12}$
             & 7    & 7     & 7 \\
    $\chi_2$ & $S^{2,2}_0 = C^1_{2} \oplus C^2_{1,7,15,27} \oplus C^3_{6,14,20} \oplus C^4_{19}$
             & 9    & 9     & 9 \\
    $\chi$   & $0 = C^0_{27} \oplus C^1_{0, 2, 26, 8} \oplus C^2_{1,13,15,27,31} \oplus C^3_{6,12,14,20} \oplus C^4_{19}$
             & 16   & 16    & 15.5 \\
    \midrule
    \multicolumn{2}{@{}l}{Approximations for \cipher{MiniMORUS1280}} \\
    \midrule
    $\chi_1$ & $S^{2,2}_0 = C^0_{51} \oplus C^1_{0, 33, 55} \oplus C^2_{4,37,46} \oplus C^3_{50}$
             & 7    & 7     & 7 \\
    $\chi_2$ & $S^{2,2}_0 = C^1_{25} \oplus C^2_{7,29,38,51} \oplus C^3_{11,20,42} \oplus C^4_{24}$
             & 9    & 9     & 9 \\
    $\chi$   & $0 = C^0_{51} \oplus C^1_{0, 25, 33, 55} \oplus C^2_{4,7,29,37,38,46,51} \oplus C^3_{11,20,42,50} \oplus C^4_{24}$
             & 16   & 16    & {??} \\
    \bottomrule
  \end{tabular}
  }
\end{table}

@@

\begin{figure}
  \substatesfalse
  % \substatesfalse to label state words and/or masks
  \centering
  \begin{subfigure}{.32\textwidth}
  \centering
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, beta]
      (C) -- node[right] {$i$} (lll-1)
      (lll-1) -- (tlll-1) (lll-1) -- (xor-1) (xor-1) -- (xnd-1)
      (tlll-1) -- (W-20) node[above] {$i$}
      (xor-1) -- (txor-1) %(txor-1) -- (W-21) node[above] {$i$}
      %(xnd-1) -- (and-1)
      (W40) node[below] {\phantom{$i$}}
      ;
    \draw[trail, gamma]
      %(W-21) node[above] {$i$}
      (txor-1) -- (xnd1)
      (xnd1) -- (xor1) (xor1) -- (lll1)
      %(xnd1) -- (and1)
      (xor1) -- (txor1)
      (txor1) -- (W-24) node[above] {$i$}
      (lll1) -- (W41) node[below] {$i+b_1$}
      ;
    %\draw[trail, beta,  dotted] (and-1) -- (and-1-|and1);
    %\draw[trail, gamma, dotted] (and1)  -- (and-1-|and1) node[above, black] {=};
    \draw[trail, beta]  (xnd-1) -- ++(-.5,0) coordinate (tmp) -- (tmp|-xnd1);
    \draw[trail, gamma] (tmp|-xnd1) node[below, black] {=} -- (xnd1);
  \end{tikzpicture}%}}}
  \caption*{\cipher{MiniMORUS}: weight 0 (not 2)} %$S^{j,1}_i, S^{j,4}_i, S^{j+1,1}_{i-1}$ ($w\!=\!1$)}
  \end{subfigure}
  \qquad
  \begin{subfigure}{.32\textwidth}
  \centering
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, beta]
      (C) -- node[right] {$i$} (lll-1)
      (lll-1) -- (tlll-1) (lll-1) -- (xor-1) (xor-1) -- (xnd-1)
      (tlll-1) -- (W-20) node[above] {$i$}
      (xor-1) -- (txor-1) %(txor-1) -- (W-21) node[above] {$i$}
      (xnd-1) -- (and-1)
      (W40) node[below] {\phantom{$i$}}
      ;
    \draw[trail, gamma]
      %(W-21) node[above] {$i$}
      (txor-1) -- (xnd1)
      (xnd1) -- (xor1) (xor1) -- (lll1)
      (xnd1) -- (and1)
      (xor1) -- (txor1)
      (txor1) -- (W-24) node[above] {$i$}
      (lll1) -- (W41) node[below] {$i+b_1$}
      ;
    \draw[trail, beta,  dashed] (and-1.east|-tanB-1) -- (tanB-1);
    \draw[trail, gamma, dashed] (tanB-1)  -- (tanB1) -- (tanB1-|and1.east);
  \end{tikzpicture}%}}}
  \caption*{\cipher{MORUS}: weight $4 \times 1$ (not $4 \times 2$)} %$S^{j,1}_i, S^{j,4}_i, S^{j+1,1}_{i-1}$ ($w\!=\!1$)}
  \end{subfigure}
  \caption{Weight of $\beta^t_i \oplus \gamma^t_i$ for \cipher{MiniMORUS} and \cipher{MORUS}.}
  \label{fig:trailcollision}
\end{figure}

>>>>

\usepackage{tikz}
\usetikzlibrary{calc,math,arrows.meta}

@@

In order to build the equation for the full trail, we start with $E^2_0$:
\[
S^2_{2,0} \oplus S^{3}_{2,b_2} \oplus S^{3}_{0,0} = S^2_{3,0} \cdot S^2_{4,0}.
\]
In order to cancel the $S^{3}_{0,0}$ term on the left-hand side, we add in the equation $A^2_{-b_0}$ (where the sum of two equations of the form $a = b$ and $c = d$ is defined to be $a+c = b+d$). This yields:
\begin{align*}
&S^2_{2,0} \oplus S^3_{2,b_2} \oplus C^2_{-b_0}\\
=\; &S^2_{3,0} \cdot S^2_{4,0} \oplus S^2_{1,-b_0} \oplus S^2_{3,-b_0} \oplus S^2_{1,-b_0} \cdot S^2_{2,-b_0} \oplus S^2_{2,-b_0} \cdot S^2_{3,-b_0}.
\end{align*}
We then need to cancel two terms of the form $S^t_{2,i}$. To do this, we add in the equations $D^t_i$ for appropriate choices of $t$ and $i$. This replaces the two $S^t_{2,i}$ terms by four $S^t_{4,i}$ terms. By using equation $B^t_i$ four times, we can then replace these four $S^t_{4,i}$ terms by eight $S^t_{1,i}$ terms. By applying equation $B^t_i$ eight times, these eight $S^t_{1,i}$ terms can in turn be replaced by eight $S^t_{0,i}$ terms (and some ciphertext terms). Finally, applying $A^t_i$ eight times allows us to replace these eight $S^t_{0,i}$ terms by only ciphertext bits. Ultimately, for \cipher{MiniMORUS640}, this yields the equation:

@@

The equation for \cipher{MiniMORUS1280} is very similar, and is given in \Cref{sec:traileq}.

@@

In the equation for \cipher{MiniMORUS640} from the previous section, each line on the right-hand side of the equality involves distinct $S^t_{i,j}$ terms, and each line has a weight of 1. By the piling-up lemma, it follows that if we assume distinct $S^t_{i,j}$ terms to be uniform and independent, then the expression on the right-hand side has a weight of 16. Hence the linear combination of ciphertext bits on the left-hand side has an expected bias of $2^{-16}$. The same holds for \cipher{MiniMORUS1280} (cf. \Cref{sec:traileq}).

@@

\begin{table}[h!]
  \caption{Experimental verification of trail biases.}
  \label{tab:miniapproximations}
  \centerline{
  \begin{tabular}{@{}llSSS@{}}
    \toprule
    & & \multicolumn{3}{@{}c@{}}{Weight} \\
    \cmidrule{3-5}
    \multicolumn{2}{@{}l}{Approximations for \cipher{MiniMORUS640}}
             & {Exp.} & {Bool.} & {Meas.} \\
    \midrule
    $\chi_1$ & $S^{2,2}_0 = C^0_{27} \oplus C^1_{0, 8, 26} \oplus C^2_{7,13,31} \oplus C^3_{12}$
             & 7    & 7     & 7 \\
    $\chi_2$ & $S^{2,2}_0 = C^1_{2} \oplus C^2_{1,7,15,27} \oplus C^3_{6,14,20} \oplus C^4_{19}$
             & 9    & 9     & 9 \\
    $\chi$   & $0 = C^0_{27} \oplus C^1_{0, 2, 26, 8} \oplus C^2_{1,13,15,27,31} \oplus C^3_{6,12,14,20} \oplus C^4_{19}$
             & 16   & 16    & 15.5 \\
    \midrule
    \multicolumn{2}{@{}l}{Approximations for \cipher{MiniMORUS1280}} \\
    \midrule
    $\chi_1$ & $S^{2,2}_0 = C^0_{51} \oplus C^1_{0, 33, 55} \oplus C^2_{4,37,46} \oplus C^3_{50}$
             & 7    & 7     & 7 \\
    $\chi_2$ & $S^{2,2}_0 = C^1_{25} \oplus C^2_{7,29,38,51} \oplus C^3_{11,20,42} \oplus C^4_{24}$
             & 9    & 9     & 9 \\
    $\chi$   & $0 = C^0_{51} \oplus C^1_{0, 25, 33, 55} \oplus C^2_{4,7,29,37,38,46,51} \oplus C^3_{11,20,42,50} \oplus C^4_{24}$
             & 16   & 16    & 15.9 \\
    \bottomrule
  \end{tabular}
  }
\end{table}

@@

\begin{figure}
  \substatesfalse
  % \substatesfalse to label state words and/or masks
  \centering
  \begin{subfigure}{.32\textwidth}
  \centering
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, beta]
      (C) -- node[right] {$i$} (lll-1)
      (lll-1) -- (tlll-1) (lll-1) -- (xor-1) (xor-1) -- (xnd-1)
      (tlll-1) -- (W-20) node[above] {$i$}
      (xor-1) -- (txor-1) %(txor-1) -- (W-21) node[above] {$i$}
      %(xnd-1) -- (and-1)
      (W40) node[below] {\phantom{$i$}}
      ;
    \draw[trail, gamma]
      %(W-21) node[above] {$i$}
      (txor-1) -- (xnd1)
      (xnd1) -- (xor1) (xor1) -- (lll1)
      %(xnd1) -- (and1)
      (xor1) -- (txor1)
      (txor1) -- (W-24) node[above] {$i$}
      (lll1) -- (W41) node[below] {$i+b_1$}
      ;
    %\draw[trail, beta,  dotted] (and-1) -- (and-1-|and1);
    %\draw[trail, gamma, dotted] (and1)  -- (and-1-|and1) node[above, black] {=};
    \node (eq) at ($(and-1|-and1)+(-1,-2.4pt)$) {=};
    \draw[beta,{Circle[sep=-1.8pt]}-]  ($(and-1)+(0,-2.4pt)$) -| (eq);
    \draw[gamma,{Circle[sep=-1.8pt]}-] ($(and1)+(0,-2.4pt)$) -- (eq);
  \end{tikzpicture}%}}}
  \caption*{\cipher{MiniMORUS}: weight 0 (not 2)} %$S^{j,1}_i, S^{j,4}_i, S^{j+1,1}_{i-1}$ ($w\!=\!1$)}
  \end{subfigure}
  \qquad
  \begin{subfigure}{.32\textwidth}
  \centering
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, beta]
      (C) -- node[right] {$i$} (lll-1)
      (lll-1) -- (tlll-1) (lll-1) -- (xor-1) (xor-1) -- (xnd-1)
      (tlll-1) -- (W-20) node[above] {$i$}
      (xor-1) -- (txor-1) %(txor-1) -- (W-21) node[above] {$i$}
      (xnd-1) -- (and-1)
      (W40) node[below] {\phantom{$i$}}
      ;
    \draw[trail, gamma]
      %(W-21) node[above] {$i$}
      (txor-1) -- (xnd1)
      (xnd1) -- (xor1) (xor1) -- (lll1)
      (xnd1) -- (and1)
      (xor1) -- (txor1)
      (txor1) -- (W-24) node[above] {$i$}
      (lll1) -- (W41) node[below] {$i+b_1$}
      ;
    \draw[trail, beta,  dashed] (and-1.east|-tanB-1) -- (tanB-1);
    \draw[trail, gamma, dashed] (tanB-1)  -- (tanB1) -- (tanB1-|and1.east);
  \end{tikzpicture}%}}}
  \caption*{\cipher{MORUS}: weight $4 \times 1$ (not $4 \times 2$)} %$S^{j,1}_i, S^{j,4}_i, S^{j+1,1}_{i-1}$ ($w\!=\!1$)}
  \end{subfigure}
  \caption{Weight of $\beta^t_i \oplus \gamma^t_i$ for \cipher{MiniMORUS} and \cipher{MORUS}.}
  \label{fig:trailcollision}
\end{figure}
